Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-06 manuscript/manuscript.tex
small tweaks to first three results sections

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution. 
To ensure computational tractability, $\omega$ takes on the mean value of $K$ bins ($K=4$ shown). 
(C) Substitution models can both, one, or neither of these features. 
We used models from the GY94 and the ExpCM families. 
}
\end{figure}

It is well known that protein sites evolve under different pressures. 
Some sites appear to tolerate many amino acids while other sites appear to tolerate only a select few. 
Therefore, it is important to use substitution models which are able to describe this purifying selection, and the resulting site-to-site rate variation, when modeling protein evolution.

The standard strategy to account for purifying selection is to allow the rate of non-synonymous change to vary among sites following some statistical distribution. 
This variation is commonly implemented by allowing the rate parameter, $\omega$, to follow a $\Gamma$-distribution ($\Gamma\omega$).
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is assigned to the mean of each bin and the site likelihood is the average likelihood given each of the $\omega$ values. 
Under this model, some sites would have a higher rate of non-synonymous change than others but this rate would be agnostic to the amino-acid identities themselves. 
All non-synonymous changes would be treated as equally.

Another strategy to account for purifying selection is to define a substitution model with a site-specific stationary state. 
Such models describe the evolution at each site uniquely and independently of the other sites. 
For example, models from the Halpern-Bruno mutation-selection family explicitly describe the frequency of each amino acid at each site. 
However, these more mechanistic models, without an arbitrary statistical distribution, come at a cost in the form of an increased number of parameters. 
When using a mutation-selection model is crucial to invoke a strategy to avoiding overfitting the data. 
We chose to define the amino-acid frequency parameters \textit{a priori} using measurements from a high-throughput functional assay called deep mutational scanning, Experimentally Informed Codon Models (ExpCMs).

These two strategies, $\Gamma\omega$ rate variation and non-uniform stationary states, are not truly distinct. 
Differences in stationary state manifest as differences in non-synonymous rates across sites \citep{spielman2015relationship}. 
Moreover, these two strategies are not mutually exclusive. 
Technically, substitution models can have a non-uniform stationary state and $\Gamma\omega$ rate variation and biologically, these models can account for both strong purifying and strong positive selection.

@@

Along a single branch, a substitution model transforms sequence divergence to branch length, which under a molecular clock assumption is proportional to time. 
However, every substitution model has a stationary state, the expected sequence divergence given a very long time, at which the relationship between sequence divergence and time is lost. 
The stationary state is represented by the long ``tails" in \ref{fig:decay}, where the expected sequence divergence remains constant as time increases.

A substitution model's stationary state is defined by the parameters in the model and can vary from model to model. 
Some model features affect how long it takes a model to reach its stationary state, whereas other features affect the stationary state itself. 
Adding $\Gamma\omega$ rate variation affects the ``path" to the stationary state but does not affect the stationary state itself. 
In addition, while $\Gamma\omega$ accounts for site-to-site rate variation, the resulting stationary state is identical at each site in the protein. 
Under these models with uniform stationary states, a mutationally constrained site and a site which tolerates many mutations will have the sequence divergence given a very long time. 
Models with non-uniform stationary states, by definition, acknowledge that differences in sites will persist even at a very long period of time. 
The difference in branch length estimation will be most dramatic at constrained sites in the protein. 
At a constrained site, the ExpCM will estimate a longer branch than the GY94 model because the overall sequence divergence at that site is expected to be small.

This difference in branch length estimation cannot be recapitulated by more complex modeling of the $\omega$ parameter. 
We transformed the GY94 model into a site-specific model by inferring a unique $\omega$ value from the ExpCM, following the work of \cite{spielman2015relationship}, for each site in the protein. 
This model takes longer to reach stationary state than the GY94 model ultimately has the same divergence level at stationary state. 
As this model shows, no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state model.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{simulations}
\textbf{Model performance under simulated, site-specific data.} 
Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
The randomized ExpCMs have amino-acid profiles shuffled among the sites 
These randomized models are still site-specific but the relationship between the site and the experimental data is broken. 
Grey points represent the length of one branch and the black points are the mean branch lengths over eight simulations. 
The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
}
\end{figure}

First, we wanted to test the effect of substitution models on branch length estimation given sequences with site-specific amino-acid frequencies. 
To this end, we simulated sequences under and ExpCM and re-inferred the branch lengths using the different substitution models described in \ref{fig:model_feature}C. 
This allowed us to test whether or not a model with a uniform stationary state could accurately estimate branch lengths and, if not, whether or not the addition of $\Gamma\omega$ could prevent branch length underestimation.

As expected, the ExpCM and ExpCM+$\Gamma\omega$ estimated the branch lengths of the simulated sequences accurately. 
As the branch lengths increase, the variance in the estimation of an given branch increases between the simulations but this error is not biased towards over- or underestimation.

The models with a uniform stationary state consistently underestimate the long branches. 
This underestimation is seen with both the GY94 and the GY94+$\Gamma\omega$, indicating that accounting for rate variation via the rate parameter cannot prevent branch length underestimation at these long branches. 
The ExpCM can also be transformed into a uniform stationary state model by setting the amino-acid at every site to the average value for that amino acid across all sites. 
In contrast to the GY94 models, this average control does not have preference for amino acid and does extract some information from the experiments, but like the GY94 models the stationary state is uniform across sites. 
These models performed similarly to the GY94 and GY94+$\Gamma\omega$ and underestimate long branches.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

@@

\begin{itemize}
\item We tested the effect substitution model choice on branch length estimation using influenza HA sequences. 
\begin{itemize}
\item These sequences presumably evolve with site-specific frequencies but unlike the simulations we do not \textit{know} the model
\item These sequences are from fairly diverged proteins. Some sequences share only 40\% identity on the amino acid level 
\item The HA tree has closely related subtypes with long branches between subtypes.
\end{itemize}
\item Both $\Gamma\omega$ and non-uniform stationary states result in an extension of branch lengths. 
\begin{itemize}
\item The addition of $\Gamma\omega$ extends branches independent of preferences. 
\item The addition of preferences extends branches \textit{from the focal sequence} independent of $\Gamma\omega$. 
\end{itemize}
\item The extension of branch length from the focal sequence shows that the ExpCM does not have equal "relevance" across the whole tree. 
\begin{itemize}
\item This result is not entirely surprising. DMS measure the effects of single amino-acid changes in the context of a single genetic background. 
\item It is expected that there would be epistatic changes across such a highly diverged phylogenetic tree. ``Shifting preferences".
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item We investigated the competing effects of ``shifting preferences" and long branches by comparing ExpCMs with different "relevances"
\item We use the fit stringency parameter $\beta$ as a measure of relevance. 
\begin{itemize}
\item The stringency parameter rescales the preferences. A stringency parameter with a value greater than one indicates that selection in nature prefers the same amino acids as selection in lab but with greater strength. 
\item There is an inverse relationship between the stringency parameter and overall sequence divergence. As the overall divergence of the tree increases, the stringency parameter decreases. 
\end{itemize}
\item \skhcomment{hypothetical data} There is also a positive relationship between the stringency parameter and branch lengths 
\begin{itemize}
\item We optimized the same tree with ExpCMs with different stringency parameters. 
\item The ExpCMs with higher stringency values estimate longer branches than the ExpCMs with lower stringency values. 
\end{itemize}
\end{itemize}

@@

>>>>

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
A $\Gamma$ distribution with $K=4$ bins is shown here. 
(C) Substitution models can both, one, or neither of these features. 
We used models from the GY94 and the ExpCM families. 
}
\end{figure}

\skhcomment{Should this all be in the introduction?}

In nature, viral proteins evolve under both purifying selection, to maintain structure and function, and diversifying selection, to escape immune pressure. 
However, neither one of these constraints is homogenous across the entire protein. 
Some protein sites appear to be tolerant of many mutations, and sample many amino acids, while others appear to be mutationally tolerant, and sample only a select few. 
When modeling evolution it is important to account for this heterogeneity and phylogenetic substitution models do this in a variety of ways.

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution (\cite{yang1994maximum, yang2000codon}, \ref{fig:model_feature}B). 
Under this model, some sites would have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves and all non-synonymous changes at at given site are treated equally. 
In contrast, models from the mutation-selection family \cite{halpern1998evolutionary} account for purifying selection by explicitly defining the frequency of each amino acid at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}A).

The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While most codon models have $<$ 20 parameters, mutation-selection models will have more than 10,000 parameters for an averaged size protein. 
As with all parameter-rich models, it is important to invoke some strategy to avoid overfitting when using a mutation-selection model.
We chose to define the amino-acid frequency parameters \textit{a priori} from a high-throughput assay called deep mutational scanning. 
The resulting model, an Experimentally Informed Codon Model (ExpCM) \citep{bloom2014experimentally, hilton2017phydms}, has a parameter space comparable to the other codon models.

These two strategies to account for site-to-site rate variation are not mutually exclusive. 
While differences in stationary state manifest as differences in non-synonymous rates across sites, these models are most appropriate for modeling purifying selection \citep{spielman2015relationship}.
Proteins which are under strong purifying and diversifying selection may be better modeled by a combination of the two. 
For this study, we used four models from the GY94 and ExpCM families, which together represent all possible combinations of these two strategies (\ref{fig:model_feature}C).

@@

Along a single branch, a substitution model transforms sequence divergence into branch length. 
Under the molecular clock assumption, this branch length is proportional to time.  
However, every substitution model has a stationary state, defining the expected sequence divergence given a very long time, at which this relationship between sequence divergence and time is lost. 
In \ref{fig:decay}, the stationary state is represented by the long ``tails" where the expected sequence divergence remains constant as time increases.

The specific parametrization scheme of a model will control features of its stationary state. 
Comparing GY94 to GY94+$\Gamma\omega$ will change how long it takes for the model to reach the stationary state but it does not change the expected sequence divergence at the stationary state. 
Furthermore, the GY94 and GY94+$\Gamma\omega$ stationary state is exactly the same at each site in the protein. 
Site-specific stationary states modulate both the amount of time to reach and the overall divergence level at the stationary state. 
These differences arise from whether or not the site is modeled as constrained or mutationally tolerant.  
The stationary state is important to branch length estimation because differences in expected divergence will lead to differences in branch length estimation and branch lengths cannot be estimated with any accuracy when two sequences have reached the divergence level of the stationary state.

The difference in branch length estimation between uniform and site-specific stationary state models will be most dramatic at constrained sites. 
Most uniform stationary state models define roughly equal frequency of each amino acid, in essence they assume that each site is mutationally tolerant. 
If the sequence divergence at constrained site is quite small, a site-uniform model will expect a small amount of time to have passed, a short branch, and a site-specific model will expect that a large amount of time has passed, long branch. 
At a mutationally tolerant site, both models will estimate roughly the same branch length for any given divergence.

This difference in branch length estimation cannot be recapitulated by more complex modeling of the $\omega$ parameter. 
We transformed the GY94 model into a site-specific model by inferring a unique $\omega$ value from the ExpCM for each site in the protein, following the work of \cite{spielman2015relationship}. 
This model takes longer to reach stationary state than the GY94 model but, ultimately, has the same divergence level at stationary state. 
As this model shows, no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state model.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{simulations}
\textbf{Model performance under simulated, site-specific data.} 
Alignments were simulated under an ExpCM along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
The averaged ExpCMs amino-acid frequencies defined by the average preference of that amino acid across all sites in the protein. 
While these models extract information from the deep mutational scanning experiment, they are not site-specific. 
Grey points represent the length of one branch and the black points are the mean branch lengths over ten simulations. 
The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
}
\end{figure}

To test the effect of substitution models on branch length estimation given sequences with site-specific amino-acid frequencies we simulated sequences under an ExpCM and re-inferred the branch lengths using the different substitution models described in \ref{fig:model_feature}C.
As expected, the ExpCM and ExpCM+$\Gamma\omega$ estimated the branch lengths of the simulated sequences accurately. 
The variance in the estimation of an given branch increases between the simulations increased as branch length increased but this error is not biased towards over- or underestimation.
The models with a uniform stationary state consistently underestimate the long branches. 
This underestimation is seen with both the GY94 and the GY94+$\Gamma\omega$, indicating that accounting for rate variation via the rate parameter cannot prevent branch length underestimation at these long branches. 
The ExpCM can also be transformed into a uniform stationary state model by setting the amino-acid at every site to the average value for that amino acid across all sites. 
In contrast to the GY94 models, this average control does not have preference for amino acid and does extract some information from the experiments, but like the GY94 models the stationary state is uniform across sites. 
These models performed similarly to the GY94 and GY94+$\Gamma\omega$ and underestimated long branches.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to GY94 models.} ).
}
\end{figure}

@@

\begin{itemize}
\item We tested the effect substitution model choice on branch length estimation using influenza HA sequences. 
\begin{itemize}
\item These sequences presumably evolve with site-specific frequencies but unlike the simulations we do not \textit{know} the model
\item These sequences are from fairly diverged proteins. Some sequences share only 40\% identity on the amino acid level 
\item The HA tree has closely related subtypes with long branches between subtypes. This is good because we are looking for the effect on long branches. 
\end{itemize}
\item Both $\Gamma\omega$ and non-uniform stationary states result in an extension of branch lengths. 
\begin{itemize}
\item The addition of $\Gamma\omega$ extends branches independent of preferences. 
\item The addition of preferences extends branches \textit{from the focal sequence} independent of $\Gamma\omega$. 
\end{itemize}
\item The extension of branch length from the focal sequence shows that the ExpCM does not have equal "relevance" across the whole tree. 
\begin{itemize}
\item This result is not entirely surprising. DMS measure the effects of single amino-acid changes in the context of a single genetic background. 
\item It is expected that there would be epistatic changes across such a highly diverged phylogenetic tree. ``Shifting preferences".
\item "Sampling" from both preference sets via preference averaging extends the branch length from both focal sequences.
\end{itemize}
\end{itemize}

@@

\begin{itemize}
\item We investigated the competing effects of ``shifting preferences" and long branches by comparing ExpCMs with trees of increasing overall divergence from the focal sequence. 
\item We use the fit stringency parameter $\beta$ as a measure of relevance. 
\begin{itemize}
\item The stringency parameter rescales the preferences. A stringency parameter with a value greater than one indicates that selection in nature prefers the same amino acids as selection in lab but with greater strength. 
\item There is an inverse relationship between the stringency parameter and overall sequence divergence. As the overall divergence of the tree increases, the stringency parameter decreases. 
\end{itemize}
\item \skhcomment{hypothetical data} There is also a positive relationship between the stringency parameter and branch lengths 
\begin{itemize}
\item We optimized the same tree with ExpCMs with different stringency parameters. 
\item The ExpCMs with higher stringency values estimate longer branches than the ExpCMs with lower stringency values. 
\end{itemize}
\end{itemize}

@@

\skhcomment{Section Outline}

\begin{itemize}
\item \texttt{phylobayes} implements the mutation-selection model in a bayesian framework as a way to avoid overfitting. 
\item \texttt{phylobayes} and ExpCM (H1+H3 average) estimate branches of roughly equal length 
\begin{itemize}
\item the ExpCM estimate longer branches from the focal sequences 
\item \texttt{phylobayes} estimates longer branches from the non-focal sequences 
\end{itemize}
\end{itemize}
